SSA based global code motion of identical computations

### 2017-04-14 hoist.tex
simd benefits
Updated abstract

\title{\GVN{}-Hoist: Hoisting Computations from Branches}

@@

\maketitle
\begin{abstract}
  Code-hoisting identifies identical computations across the program and hoists
  them to a common dominator so as to save code size.  Not only code-hoisting
  saves code size, it exposes redundancies in some cases, it exposes more
  instruction level parallelism in the basic block a computation is hoisted, and
  it enables other passes like LICM to remove more redundancies.

  We present a code hoisting pass that we implemented in \LLVM{}. It is based on
  Global Value Numbering infrastructure available in \LLVM{}. The experimental
  results show an average of 2.5\% savings in code size, although the code size
  also increases in many cases because it enables more inlining. This is an
  optimistic algorithm in the sense that it considers all identical computations
  in a function as potential candidates to be hoisted. We make an extra effort
  to hoist candidates by partitioning the potential candidates in a way to
  enable partial hoisting in case common hoisting points for all the candidates
  cannot be found. We also formalize cases when register pressure will reduce as
  a result of hoisting and why sorting the list of potential candidates
  w.r.t. their DFS numbers helps hoist more candidates with less compile time
  overhead.
\end{abstract}

@@

\begin{itemize}
\item to reduce the code size of the program;
\item to improve function inlining heuristics: functions become cheaper to
  inline when their code size decreases because inliner heuristics in \LLVM{}
  depends on instruction count.
\item to expose more instruction level parallelism: by hoisting identical
  computations to be executed earlier, instruction schedulers can move heavy
  computations earlier in order to avoid pipeline bubbles;
\item to help out-of-order processors with speculative execution of branches: by
  hoisting expressions out of branches, code hoisting can effectively reduce the
  amount of code to be speculatively executed and can reduce the critical path;
\item to reduce interference/register pressure: by moving computations closer to
  the definitions of their operands in cases where hoisting would reduce the
  interference. For example if all the use operands of an expression of the form
  a = b op c are last usages (a kill) then hoisting will reduce the interference
  by two but increase by one.
\item to improve passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted and sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}

>>>>

\title{SSA based global code motion of identical computations}

@@

\maketitle
\begin{abstract}
We present a global code motion compiler optimization which schedules
identical computations across the program so as to save code size.  Not only
this code motion saves code size, it exposes redundancies in some cases, it
exposes more instruction level parallelism in the basic block when the
computations are hoisted, and it enables other passes like loop invariant
motion to remove more redundancies. The cost model to drive the code motion is
based on live range analysis on SSA representation such that the (virtual)
register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
of 2.5\% savings in code size in llvm testsuite, although the code size also
increases in many cases because it enables more inlining. We have also seen
imrpovements of approximately 5\% in a couple of SPEC 2006 benchmarks viz. gcc
and mcf, moreover, register spills reduced for almost all the SPEC 2006
bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
sense that it considers all identical computations in a function as potential
candidates. We make an extra effort to hoist candidates by partitioning the
potential candidates in a way to enable partial hoisting in case common
hoisting points for all the candidates cannot be found. We also formalize how
register pressure will reduce as a result of code-motion and why sorting the
list of potential candidates w.r.t. their depth first numbers helps hoist more
candidates with less compile time overhead.
\end{abstract}

@@

\begin{itemize}
\item to reduce the code size of the program;
\item to improve function inlining heuristics: functions become cheaper to
  inline when their code size decreases because inliner heuristics in \LLVM{}
  depends on instruction count.
\item to expose more instruction level parallelism: by hoisting identical
  computations to be executed earlier, instruction schedulers can move heavy
  computations earlier in order to avoid pipeline bubbles;
\item to help out-of-order processors with speculative execution of branches: by
  hoisting expressions out of branches, code hoisting can effectively reduce the
  amount of code to be speculatively executed and can reduce the critical path;
\item to reduce interference/register pressure: by moving computations closer to
  the definitions of their operands in cases where hoisting would reduce the
  interference. For example if all the use operands of an expression of the form
  a = b op c are last usages (a kill) then hoisting will reduce the interference
  by two but increase by one.
\item for SIMD architectures which execute all branches, code hoisting will
  reduce the total number of instructions to be executed.
\item to improve passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted and sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}
